---
title: "Basic_AOV_TreatmentAndStageDurationEffect"
author: "Shelly Trigg"
date: "4/12/2019"
output: rmarkdown::github_document
---

Load libraries
```{r, echo = FALSE}
library("MetaboAnalystR")
library("broom")
library("tidyr")
library("dplyr")
library("data.table")
library(effsize)
library(ggplot2)
library(readxl)
```

load and reformat metabolite data

```{r}
mSet<-InitDataObjects("pktable", "stat", FALSE)
mSet<-Read.TextData(mSet, "~/Desktop/UCSD-SALK-UW/NOAA/metabolomics/data/for metaboanalyst/339117_data4MetaboAnalyst.csv", "rowu", "disc")
mSet<-SanityCheckData(mSet)
mSet<-RemoveMissingPercent(mSet, percent=0.5)
mSet<-ImputeVar(mSet, method="min")
dataset <- data.frame(mSet$dataSet)
ncmpds <- length(grep("orig.", colnames(dataset)))-1

data1 <- data.frame(mSet$dataSet)
data1 <- data1[,c(6,grep("^procr.", colnames(data1)))]
data1$sample <- rownames(data1)
data1 <- data1[,c(ncol(data1),2:ncol(data1)-1)]
data1$pH <- ifelse(grepl("igh", substr(data1$cls,1,4)), "High", "Low")
data1$DO <- ifelse(grepl("igh",substr(data1$cls,6,10)), "High", "Low")
data1 <- data1[,c(1,2,654,655,3:653)]
colnames(data1) <- c("mx_sample","treatment", "pH", "DO", as.list(substring(colnames(data1[,grep("procr.",colnames(data1))]),7)))

STACKED_data <- tidyr::gather(data1, "analyte", "quant", 5:ncol(data1))


#count compounds after removing those with noise abundance
length(unique(STACKED_data$analyte))
#651

```

load lipids data

```{r}
mSet<-InitDataObjects("pktable", "stat", FALSE)
mSet<-Read.TextData(mSet, "~/Desktop/UCSD-SALK-UW/NOAA/metabolomics/ARTool_analysis/GenLipidsAll.csv", "rowu", "disc")
mSet<-SanityCheckData(mSet)
mSet<-RemoveMissingPercent(mSet, percent=0.5)
mSet<-ImputeVar(mSet, method="min")
dataset <- data.frame(mSet$dataSet)
ncmpds <- length(grep("orig.", colnames(dataset)))-1

data2 <- data.frame(mSet$dataSet)
data2 <- data2[,c(6,grep("^procr.", colnames(data2)))]
data2$sample <- rownames(data2)
data2 <- data2[,c(ncol(data2),2:ncol(data2)-1)]
data2$pH <- ifelse(grepl("igh", substr(data2$cls,1,4)), "High", "Low")
data2$DO <- ifelse(grepl("igh",substr(data2$cls,6,10)), "High", "Low")
ncol(data2)
data2 <- data2[,c(1,2,3099, 3100, 3:3098)]
colnames(data2) <- c("mx_sample","treatment", "pH", "DO", as.list(substring(colnames(data2[,grep("procr.",colnames(data2))]),7)))

STACKED_data2 <- tidyr::gather(data2, "analyte", "quant", 5:ncol(data2))

#count compounds after removing those with noise abundance
length(unique(STACKED_data2$analyte))
#3096

```


read in stage duration data
```{r}
duration <- read.csv("~/Desktop/UCSD-SALK-UW/NOAA/metabolomics/data/crab_duration_info.csv", stringsAsFactors = FALSE)
```

read in sample ID info data
```{r}
library(readxl)
sampleidinfo <- read_xlsx("~/Desktop/UCSD-SALK-UW/NOAA/metabolomics/data/SampleID_CrabID_info.xlsx")
```

merge sample ID info and stage duration info with compound abundance data
```{r}
colnames(STACKED_data)[1] <- "Sample.Number"

STACKED_data <- merge(STACKED_data,sampleidinfo, by = "Sample.Number")
STACKED_data <- merge(STACKED_data,duration, by = "crabID")

colnames(STACKED_data2)[1] <- "mx.sample"
STACKED_data2 <- merge(STACKED_data2,sampleidinfo, by = "mx.sample")
STACKED_data2 <- merge(STACKED_data2,duration, by = "crabID")
```

## STATISTICS

### Check if there are effects from stage duration or days on treatment

test for megalopae duration effect on metabolite quant
```{r}
fit <- aov(quant ~ M, data = STACKED_data)
summary(fit)
```

test for J1 duration effect on metabolite quant
```{r}
fit <- aov(quant ~ J1, data = STACKED_data)
summary(fit)
```

test for MtoJ2 duration effect on metabolite quant
```{r}
fit <- aov(quant ~ MtoJ2, data = STACKED_data)
summary(fit)
```

test for startTOfreeze duration effect on quant
```{r}
fit <- aov(quant ~ startTOfreeze, data = STACKED_data)
summary(fit)
```

test for treatment effect on quant
```{r}
fit <- aov(quant ~ treatment.x, data = STACKED_data)
summary(fit)
```
 

### Analysis for lipid data

test for Megalopae duration effect on lipid quant
```{r}
fit <- aov(quant ~ M, data = STACKED_data2)
summary(fit)
```   

test for J1 duration effect on lipid quant
```{r}
fit <- aov(quant ~ J1, data = STACKED_data2)
summary(fit)
```    

test for MtoJ2 duration effect on quant
```{r}
fit <- aov(quant ~ MtoJ2, data = STACKED_data2)
summary(fit)
``` 

test for startTOfreeze duration (days on experiment) effect on quant
```{r}
fit <- aov(quant ~ startTOfreeze, data = STACKED_data2)
summary(fit)
```

test for treatment effect on quant
```{r}
fit <- aov(quant ~ treatment.x, data = STACKED_data2)
summary(fit)
```
